---
title: "Getting Started with seuratPassport"
author:
  - name: Sedat Kacar
    affiliation: Indiana University
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Getting Started with seuratPassport}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  eval      = FALSE
)
```

# Introduction

`seuratPassport` solves a common problem in single-cell RNA-seq analysis:
**Seurat objects accumulate complex processing histories, but that context
disappears once collaborators open the `.rds` file.**

This package embeds a persistent metadata *passport* directly inside
`@misc$passport` of every Seurat object. Because the passport travels inside
the `.rds` file, there are no external spreadsheets to lose track of.

The package provides three functions:

| Function | Purpose |
|---|---|
| `seuratPassport()` | Open the Shiny popup to fill or update a passport |
| `read_passport()` | Print the full passport and processing log to console |
| `log_step()` | Append a processing step to the object's log |

# Installation

```{r install, eval=FALSE}
# From Bioconductor (once accepted):
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("seuratPassport")

# Development version from GitHub:
# devtools::install_github("sedatkacar56/seuratPassport")
```

# Quick Start

## Stamping a root object

Call `seuratPassport()` on any Seurat object. An interactive Shiny popup
opens in your RStudio Viewer. Fill in the fields and click **Done**.

```{r stamp-root}
library(seuratPassport)

# WTHeme is your Seurat object
WTHeme <- seuratPassport(WTHeme)
```

The passport is now stored in `WTHeme@misc$passport` and will persist
whenever you `saveRDS(WTHeme, "WTHeme.rds")`.

## Stamping a child object (subsetting workflow)

When you subset an object, pass the parent so the lineage is filled
automatically — no typing needed.

```{r stamp-child}
# Subset endothelial cells from WTHeme
EndofrHeme <- subset(WTHeme, idents = "Endothelial")

# parent = WTHeme fills parent_id and lineage automatically
EndofrHeme <- seuratPassport(EndofrHeme, parent = WTHeme)
```

`EndofrHeme@misc$passport$lineage` will now contain the full ancestry chain
back to the root object.

## Reading a passport

```{r read}
# Print to console
read_passport(WTHeme)

# Or via the main function with read = TRUE
seuratPassport(WTHeme, read = TRUE)
```

Example output:

```
========== PASSPORT ==========
Object ID  : WTHeme
RDS Self   : 224
Created    : 2026-03-19 05:00:00
-------- Animal --------
Animal ID  : M01
Species    : Mus musculus
Sex        : male
Age        : P60
Condition  : control
Tissue     : lung
-------- Experiment --------
Project    : memory_study
Researcher : Sedat Kacar
Date       : 2026-03-19
Notes      : First sequencing run
-------- Lineage --------
Parent     : root
RDS Parent : root
Chain      : root
Children   : EndofrHeme224, gCapC
RDS Children: 225, 226
======= PROCESSING LOG =======
No processing steps logged yet.
==============================
```

# Logging Processing Steps

Use `log_step()` after each major processing step. This creates an auditable
record of what was done to the object, with timestamps and cell counts.

```{r log}
library(Seurat)

WTHeme <- NormalizeData(WTHeme)
WTHeme <- log_step(WTHeme, "NormalizeData",
                   params = list(method = "LogNormalize",
                                 scale_factor = 10000))

WTHeme <- FindVariableFeatures(WTHeme)
WTHeme <- log_step(WTHeme, "FindVariableFeatures",
                   params = list(nfeatures = 2000))

WTHeme <- ScaleData(WTHeme)
WTHeme <- log_step(WTHeme, "ScaleData")

WTHeme <- RunPCA(WTHeme, npcs = 30)
WTHeme <- log_step(WTHeme, "RunPCA", params = list(npcs = 30))
```

The processing log is visible at the bottom of `read_passport()` output:

```
======= PROCESSING LOG =======
[1] NormalizeData             | 8423 cells | 2026-03-19 05:01:00
[2] FindVariableFeatures      | 8423 cells | 2026-03-19 05:01:05
[3] ScaleData                 | 8423 cells | 2026-03-19 05:01:20
[4] RunPCA                    | 8423 cells | 2026-03-19 05:01:45
==============================
```

# Custom Fields

The Shiny popup includes a **Custom Fields** section where you can add
any extra key-value pairs. These are stored alongside the standard fields
and printed by `read_passport()`.

For example, you might add:
- `sequencing_platform = 10x Chromium v3`
- `genome_build = mm10`
- `batch = batch_01`

Custom fields from a previous passport are automatically pre-loaded when
you re-open the popup, so you only update what changed.

# Passport Fields Reference

## Identity

| Field | Description | Example |
|---|---|---|
| `object_id` | Name/ID of this object | `"WTHeme"` |
| `rds_self` | RDS registry number | `"224"` |
| `created` | Timestamp of stamping | auto-filled |

## Animal Info

| Field | Description | Example |
|---|---|---|
| `animal_id` | Individual animal ID | `"M01"` |
| `species` | Species | `"Mus musculus"` |
| `sex` | Sex | `"male"` |
| `age` | Age | `"P60"` |
| `condition` | Experimental condition | `"control"` |
| `tissue` | Tissue of origin | `"lung"` |

## Experiment Info

| Field | Description | Example |
|---|---|---|
| `project` | Project name | `"memory_study"` |
| `researcher` | Researcher name | `"Sedat Kacar"` |
| `date` | Experiment date | `"2026-03-19"` |
| `notes` | Free-text notes | any string |

## Lineage

| Field | Description |
|---|---|
| `parent_id` | Object ID of the direct parent (or `"root"`) |
| `rds_parent` | RDS number of the parent (or `"root"`) |
| `lineage` | Full ancestry chain as a character vector |
| `children` | Object IDs of children subset from this object |
| `rds_children` | RDS numbers of children |

# Session Info

```{r session}
sessionInfo()
```
